---
title: "Chronology and regional settlement in the Zagros, 20,000–6000 BP"
output: html_notebook
---

```{r deps}
library("tidyverse")
library("magrittr") # For T-pipe
library("sf")
library("stratigraphr")
library("rcarbon")
library("patchwork")
library("ggpubr")
```

Research compendium containing data and source code for a meta-analysis of chronology and regional settlement in the Epipalaeolithic-Neolithic of the Zagros region (20000-6000 BP).

## Materials & Methods

### Site location data

```{r data-map}
sw_asia <- st_read("data/regions/swasia.geojson")
  
rbind(
  st_read("data/regions/cyprus.geojson"),
  st_read("data/regions/canatolia.geojson"),
  st_read("data/regions/slevant.geojson"),
  st_read("data/regions/east_fcrescent.geojson"),
  st_read("data/regions/nlevant_seanatolia.geojson"),
  st_read("data/regions/desert.geojson")
) %>% 
  add_column(region = c("cyprus", "canatolia", "slevant", "east_fcrescent",
                        "nlevant_seanatolia", "desert"), .before = 1) ->
  regions
```

```{r data-site}
sites <- read_csv("./data/zagros_site.csv")
```

### Relative chronological data



### Radiocarbon data

Import and clean data.
Exclude obvious outliers and dates with unknown errors.
Assign dummy values to dates without a Lab ID.


```{r data-zagros-c14}
read_csv("./data/zagros_c14.csv") %T>% 
  {message("Imported ", nrow(.), " dates.")} %>%
  filter(!outlier) %>% 
  drop_na(c14_error) %>% 
  mutate(lab_id = if_else(is.na(lab_id), paste0("Unknown-", seq_along(lab_id)), lab_id)) %T>% 
  {message(nrow(.), " dates remaining after cleaning.")} ->
  dates

dates <- right_join(sites, dates, by = "site_id", na_matches = "never")
```

And the NERD database for cross-regional comparisons:

```{r data-nerd}
nerd <- c14bazAAR::get_c14data("nerd")
```

We calibrate them with IntCal20 and `rcarbon`.
Normalising the dates at this point is undesirable because it exaggerates calibration curve artifacts (Roberts et al. 2017).

```{r calibrate, echo=FALSE}
dates <- mutate(dates, cal = c14_calibrate(c14_age, c14_error, normalised = FALSE))
```

## Results

### Scientometrics

It's often said that the E. Fertile Crescent has "lagged behind" other parts of Southwest Asia in terms of primary field research.
We can get a rough measure of this by looking at the earliest publication date associated with each site and radiocarbon date in our database.

```{r data-pub-year, warning=FALSE}
# TODO: Deal with `Inf` warning
earliest <- function(x) {
  str_extract_all(x, r"(\d\d\d\d)", simplify = TRUE) |>
    as.numeric() |>
    min(na.rm = FALSE)
}

sites <- mutate(sites, first_pub = map_dbl(references, earliest))
dates <- mutate(dates, first_pub = map_dbl(paste(source, additional_references),
                                           earliest))
```

```{r plot-pub-year}
sites |>
  drop_na(first_pub) |>
  count(first_pub) |>
  filter(is.finite(first_pub)) |>
  mutate(ncum = cumsum(n)) |>
  ggplot(aes(first_pub, ncum)) +
  geom_area() +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10), limits = c(1950, 2021)) +
  labs(
    x = NULL,
    y = "Sites"
  ) ->
  p_pub_year1

dates |>
  drop_na(first_pub) |>
  count(first_pub) |>
  filter(is.finite(first_pub)) |>
  mutate(ncum = cumsum(n)) |>
  ggplot(aes(first_pub, ncum)) +
  geom_area() +
  scale_x_continuous(breaks = seq(1950, 2020, by = 10), limits = c(1950, 2021)) +
  labs(x = NULL,
       y = "Dates") ->
  p_pub_year2

(p_pub_year1 / p_pub_year2) +
  plot_annotation(
    title = "Cumulative number of published sites and radiocarbon dates",
    subtitle = "Eastern Fertile Crescent"
  )

# ggsave("figures/pub_year.png", width = 254, height = 143, units = "mm")
```

<!-- 
TODO: Some sort of map visualisation (animated?), after coordinates joined to dates
-->

Interestingly I'm not sure that the quantitative data bears this out: there are some big "jumps" in the publication of sites, corresponding to major surveys, but overall there looks to have been a steady stream of new data coming in since at least the late 1960s.

Using Palmisiano et al.'s NERD database, we can also compare this to other regions (this time including dates up to 2000 BP):

```{r data-nerd-pub-date, warning=FALSE}
nerd <- mutate(nerd, first_pub = map_dbl(shortref, earliest))
```

```{r plot-pub-year-byregion}
nerd |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_join(regions) |>
  drop_na(region) |>
  mutate(region = recode(region,
                         "canatolia" = "Central Anatolia",
                         "cyprus" = "Cyprus",
                         "desert" = "Arid interior",
                         "east_fcrescent" = "Eastern Fertile Crescent",
                         "nlevant_seanatolia" = "Northern Levant/Southeast Anatolia",
                         "slevant" = "Southern Levant")) |>
  drop_na(first_pub) |>
  count(region, first_pub) |>
  filter(is.finite(first_pub)) |>
  group_by(region) |>
  mutate(ncum = cumsum(n)) |>
  ggplot(aes(first_pub, ncum)) +
  facet_wrap(vars(region)) +
  geom_area() +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10), limits = c(1960, 2021)) +
  labs(title = "Cumulative number of radiocarbon dates by region",
       x = NULL,
       y = "Radiocarbon dates",
       caption = "Data: Near East Radiocarbon Dates <https://github.com/apalmisano82/NERD>")

# ggsave("figures/pub_year_byregion.png", width = 254, height = 143, units = "mm")
```

Again, the quantitative data doesn't match the "lagging behind" perception – at least in terms of radiocarbon dates.
The growth rate and total sample from the E. Fertile Crescent are roughly the same as for comparable regions;
the real exception is the explosion of high resolution site-dating in the Southern levant in the last ten years.

### Chronology

```{r plot-dates-by-site}
dates |>
  count(site_name) |>
  ggplot(aes(fct_reorder(site_name, n, .desc = TRUE), n)) +
  geom_col() +
  labs(x = NULL, y = "Number of dates") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 6))

#ggsave("figures/dates_by_site.png", width = 254, height = 143, units = "mm")
```

```{r plot-period-summary}
dates |>
  pivot_longer(period_LP:period_H, names_to = "period", values_to = "certainty",
               values_drop_na = TRUE) |>
  mutate(
    period = str_remove(period, "period_"),
    period = ordered(period,
                     levels = c("UP",
                                "MP",
                                "EP",
                                "N0",
                                "N1.1",
                                "N1.2",
                                "N2",
                                "N3",
                                "C1",
                                "C2",
                                "C3",
                                "B",
                                "H"
                                ))
  ) |>
  count(period, site_id) |>
  group_by(period) |>
  summarise(n_sites = n(), n_dates = sum(n)) |>
  ggplot(aes(x = period, y = n_sites)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Number of sites")

# ggsave("figures/periods_sites.png", width = 127, height = 143, units = "mm")

dates |>
  pivot_longer(period_LP:period_H, names_to = "period", values_to = "certainty",
               values_drop_na = TRUE) |>
  mutate(
    period = str_remove(period, "period_"),
    period = ordered(period,
                     levels = c("UP",
                                "MP",
                                "EP",
                                "N0",
                                "N1.1",
                                "N1.2",
                                "N2",
                                "N3",
                                "C1",
                                "C2",
                                "C3",
                                "B",
                                "H"
                                ))
  ) |>
  count(period, site_id) |>
  group_by(period) |>
  summarise(n_sites = n(), n_dates = sum(n)) |>
  ggplot(aes(x = period, y = n_dates)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Number of dates")

# ggsave("figures/periods_dates.png", width = 127, height = 143, units = "mm")
```

Plotting all dates for which relative chronological information is available:

```{r data-sum-periods, warning=FALSE}
dates |>
  pivot_longer(period_LP:period_H, names_to = "period", values_to = "certainty",
               values_drop_na = TRUE) |>
  mutate(
    period = str_remove(period, "period_"),
    period = ordered(period,
                     levels = c("UP",
                                "MP",
                                "EP",
                                "N0",
                                "N1.1",
                                "N1.2",
                                "N2",
                                "N3",
                                "C1",
                                "C2",
                                "C3",
                                "B",
                                "H"
                                ))
  ) |>
  group_by(period) |>
  summarise(cal_sum = c14_sum(cal)) ->
  spd_byperiod_all

dates |>
  pivot_longer(period_LP:period_H, names_to = "period", values_to = "certainty",
               values_drop_na = TRUE) |>
  filter(certainty == "x") |>
  mutate(
    period = str_remove(period, "period_"),
    period = ordered(period,
                     levels = c("UP",
                                "MP",
                                "EP",
                                "N0",
                                "N1.1",
                                "N1.2",
                                "N2",
                                "N3",
                                "C1",
                                "C2",
                                "C3",
                                "B",
                                "H"
                                ))
  ) |>
  group_by(period) |>
  summarise(cal_sum = c14_sum(cal)) ->
  spd_byperiod_strict
```

```{r plot-periods}
spd_byperiod_all %>% 
  filter(period != "MP") %>% 
  unnest(cal_sum) %>% 
  ggplot(aes(calBP, PrDens)) +
  facet_wrap(vars(period), nrow = 1, scales = "free_x") +
  geom_area() +
  scale_x_reverse(limits = c(20000, 6000), 
                  breaks = seq(20000, 6000, by = -2000)) +  coord_flip() +
  labs(title = "Summed radiocarbon by typochronological unit",
       subtitle = "All sites with typochronological data available",
       x = "cal BP", y = NULL) + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank())

# ggsave("figures/c14_periods_all.png", width = 254, height = 143, units = "mm")

spd_byperiod_strict %>% 
  filter(period != "MP") %>% 
  unnest(cal_sum) %>% 
  ggplot(aes(calBP, PrDens)) +
  facet_wrap(vars(period), nrow = 1, scales = "free_x") +
  geom_area() +
  scale_x_reverse(limits = c(20000, 6000), 
                  breaks = seq(20000, 6000, by = -2000)) +
  coord_flip() +
  labs(title = "Summed radiocarbon by typochronological unit",
       subtitle = "Only sites with secure typochronological data available",
       x = "cal BP", y = NULL) + 
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank())

# ggsave("figures/c14_periods_strict.png", width = 254, height = 143, units = "mm")
```

### Summed probability distribution (SPD)

```{r echo=FALSE}
simple_spd <- c14_sum(dates$cal,
                      datenormalised = FALSE,
                      spdnormalised = TRUE)

ggplot() +
  geom_area(data=simple_spd[[1]], mapping=aes(x=calBP, y=PrDens)) +
  scale_x_reverse() +
  theme_pubclean()
```

The simple SPD is definitely bullshit, because certain sites have way more dates than others.

```{r}
ggplot(dates, aes(x = site_name)) +
  geom_bar(stat = "count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.3))
```

Binning dates by site reduces this effect.

```{r echo=FALSE}
# TODO: refresh old code
library("rcarbon")
dates2 <- filter(dates, c14_age > 5100 & c14_age < 24000)
cal_dates <- calibrate(dates2$c14_age, dates2$c14_error, 
                       # ids = dates2$lab_id,
                       timeRange = c(20000, 6000),
                       normalised = FALSE)

bsites_spd <- spd(cal_dates, timeRange = c(20000, 6000), datenormalised = FALSE,
                  bins = dates$site_id, spdnormalised = TRUE)

ggplot() +
  # geom_area(data=simple_spd$grid, mapping=aes(x=calBP, y=PrDens), fill="grey") +
  geom_area(data=bsites_spd$grid, mapping=aes(x=calBP, y=PrDens)) +
  scale_x_reverse()
```

Or, we can try Roberts et al.'s (2017) method of creating arbitrary bins consisting of any dates from the same site within a given interval (*h*) in uncalibrated years. However, this method may be sensitive to the choice of *h*.

```{r echo=FALSE}
bintrv <- lapply(c(10, 25, 50, 100, 200, 500), binPrep,
                 sites = dates2$site_id, ages = dates2$c14_age)
bintrv_spd <- lapply(bintrv, function(b) {
  spd(cal_dates, timeRange = c(20000, 6000), datenormalised = FALSE,
                  bins = b, spdnormalised = TRUE)
})

ggplot() +
  geom_area(data=bsites_spd$grid, mapping=aes(x=calBP, y=PrDens), fill="grey") +
  geom_line(data=bintrv_spd[[1]]$grid, mapping=aes(x=calBP, y=PrDens), colour="red") +
  geom_line(data=bintrv_spd[[2]]$grid, mapping=aes(x=calBP, y=PrDens), colour="orange") +
  geom_line(data=bintrv_spd[[3]]$grid, mapping=aes(x=calBP, y=PrDens), colour="yellow") +
  geom_line(data=bintrv_spd[[4]]$grid, mapping=aes(x=calBP, y=PrDens), colour="green") +
  geom_line(data=bintrv_spd[[5]]$grid, mapping=aes(x=calBP, y=PrDens), colour="blue") +
  geom_line(data=bintrv_spd[[6]]$grid, mapping=aes(x=calBP, y=PrDens), colour="black") +
  scale_x_reverse()
```

The sensitivity analysis shows that for values of *h* between 10 and 500, the resulting SPDs are more similar to each other than to the binned-by-site SPD. Low and high intervals appear to mostly amplify the extremes of the SPD. A middle value of *h*=200 therefore seems best.

```{r echo=FALSE}
bins <- binPrep(dates2$site_id, dates2$c14_age, 200)
zagros_spd <- spd(cal_dates, timeRange = c(20000, 6000), datenormalised = FALSE,
                  bins = bins, spdnormalised = TRUE)

ggplot() +
  geom_area(data=zagros_spd$grid, mapping=aes(x=calBP, y=PrDens)) +
  scale_x_reverse()
```

```{r plot-spd}
ggplot(zagros_spd$grid, aes(x=calBP, y=PrDens)) +
  geom_area() +
  scale_x_reverse(limits = c(20000, 6000), breaks = seq(20000, 6000, by = -2000)) +
  labs(title = "Eastern Fertile Crescent, summed probability distribution",
       subtitle = "N=933, sites=107; binned by site and 200 yr interval",
       y = NULL, x = "cal BP") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

ggsave("figures/zagros_spd.png", width = 254, height = 143, units = "mm")
```

#### Simulating a confidence envelope

Using Shennan et al.'s (2013) method, we can use Monte Carlo simulation to account for:
1. Taphonomic effects, by using an exponential null model
2. The effect of the calibration curve (simulated dates are calibrated using the same curve)

The choice of null model here is tricky. 
Uniform is clearly inappropriate because on this period we would expect both long-term population growth and taphonomic loss to be evident.
Shennan argues that exponential (the rcarbon default) should reflect these trends for large (i.e. continental?) regions over long time periods.
But here we are dealing with quite a small region, and for much of the period (c. 18000–12000 BP) the theoretical expectation is
So would a linear or logistic model be better?
Palmisiano et al. use a logistic model for even 12000–2000 BP, justifying it theoretically on the basis of the 'carrying capacity' argument and empirically on the observed shape of their curve.
Nielsen et al. also use it because they're specifically investigating a Neolithic demographic transition, as does the NDT literature in general.
I haven't been able to find any example of linear models being used.

```{r}
# TODO: Look at how Alessio did it
radiocarbon_model <- glm(PrDens ~ calBP, "binomial", zagros_spd$grid,
                         subset = calBP > 6000 & calBP < 20000)

# Use nsim>=1000 for final results
modelTest(cal_dates, dates2$c14_error, nsim = 1000, bins = bins,
          timeRange = c(20000, 6000), model = "custom", 
          predgrid = radiocarbon_model$model,
          datenormalised = FALSE, spdnormalised = TRUE, ncores=8) ->
  radiocarbon_mctest2

summary(radiocarbon_mctest)
plot(radiocarbon_mctest2)

# png("analysis/figures/spd_model_exp.png", width = 254, height = 71, units = "mm",
#     res = 300)
# plot(radiocarbon_mctest)
# dev.off()
```

### Settlement patterns
